### 2013-09-28 Publications/2013-ISBI14/01-abstract.tex
First review after Manu's comments

\begin{abstract}

Connectivity analysis on \gls*{dmri} data of the whole-brain
suffers from distortions originated by the standard 
\gls*{epi} sequences used in acquisition. These artifacts
are present at interfaces between tissues, due to the inhomogeneity
of the field derived from the discontinuity of magnetic
susceptibility. As a result, registered signal shows characteristic
geometrical deformations and signal destruction. These artifacts
suppose a significant pitfall that limits the success of tractography 
algorithms.

Several retrospective correction techniques are readily
available. In this work, we use a renowned digital phantom 
designed for evaluating tractography. We subject the phantom 
to a ``theoretically correct'' and plausible deformation that
resembles the artifact under investigation. We correct data back, 
with three standard methodologies (namely fieldmap-based, 
reversed encoding-based, and registration-based). 
Finally, we rank the methods based on their geometrical 
accuracy, the dropout compensation, and their impact on 
the resulting tractography.

\end{abstract}

>>>>

\begin{abstract}
Connectivity analysis on diffusion MRI data of the whole-brain
suffers from distortions originated by the standard 
echo-planar imaging sequences used in acquisition. %
%These artifacts are present at interfaces between tissues,
%due to the inhomogeneity of the field derived from the 
% discontinuity of magnetic susceptibility.
These images show characteristic geometrical deformations
and signal destruction that suppose a significant pitfall
limiting the success of tractography algorithms.

Several retrospective correction techniques are readily
available. In this work, we use a digital phantom designed
for the evaluation of connectivity pipelines. We subject
the phantom to a ``theoretically correct'' and plausible
deformation that resembles the artifact under investigation.
We correct data back, with three standard methodologies
(namely fieldmap-based, reversed encoding-based, and
registration-based). Finally, we rank the methods based
on their geometrical accuracy, the dropout compensation,
and their impact on the resulting tractography.
\end{abstract}%
\begin{keywords}
susceptibility artifacts, diffusion MRI, tractography, connectivity.
\end{keywords}

### 2013-09-28 Publications/2013-ISBI14/02-introduction.tex
First review after Manu's comments

However, such a complex workflow is prone to
flaws that have been detailedly reported 
\cite{jones_twenty-five_2010}.
In order to enable reliable and robust connectome
analyses of the brain, some weak processing points must
be improved. One prominent source
of inaccuracy are the susceptibility-derived artifacts,
for which \Gls*{epi} schemes are specially sensitive.
As susceptibility changes at tissue interfaces,
so does the magnetic field. The consequence of the 
inhomogeneity of the field translates in a highly 
distorted imaged anatomy and a significant signal 
destruction of certain regions of the brain 
(e.g. the orbitofrontal lobe, for the proximity of the
air surrounding sinuses). This artifact has been
profoundly described, generally within \gls*{fmri}
literature because they are usually acquired with 
\gls*{epi} as \gls*{dmri}.

First approaches to
retrospective correction \cite{jezzard_correction_1995},
relied on an extra \gls*{mri} acquisition (so-called
field mapping), that probes the inhomogeneity of the field.
A second theory-based breed of methodologies acquire a 
map of the \gls*{psf} of the \gls*{epi} readouts to correct 
for the artifact \cite{robson_measurement_1997}. 
Next generation of methodologies \cite{cordes_geometric_2000,
chiou_simple_2000}, make use of the acquisition of 
extra \gls*{epi} volumes 
with specific differences that enable correction, for instance
the opposed direction of the phase encoding gradients 
Finally, the last family of methodologies acquire an 
extra \gls*{t2} image, and use its anatomical correctness 
as reference to find the deformation map through nonlinear 
registration \cite{kybic_unwarping_2000,studholme_accurate_2000}.
Registration-based methods usually map
the \gls*{t2} image to the so-called baseline (B0) volume 
of \gls*{dmri}. The choice of \gls*{t2} is due to the strong
similarity of intensity distribution with the B0 of \gls*{dmri}
datasets. More recent works report extensions of the existing
techniques \cite{andersson_how_2003,holland_efficient_2010,
andersson_comprehensive_2012}, and combined approaches
\cite{zaitsev_point_2004}.

Even though aforementioned techniques for distortion correction
have been studied \cite{zeng_image_2002,wu_comparison_2008},
the lack of a gold-standard limits the possible benchmarking 
strategies. \cite{irfanoglu_effects_2012} raised the question
of distortion-derived impacts in tractography.
In this work, we propose an evaluation framework
using a detailed digital phantom designed for tractography.
Using this framework,
we are able to compare several correction techniques and we
provide results related to their performance in the task
itself and their impact on subsequent tractography.

>>>>

Among all the difficulties that such a complex workflow
raises \cite{jones_twenty-five_2010}, we will
address here the susceptibility-derived artifacts,
for which \Gls*{epi} schemes are specially sensitive.
As susceptibility changes at tissue interfaces,
so does the magnetic field. This inhomogeneities 
of the field translate in a highly distorted imaged 
anatomy and a significant signal  destruction of 
certain regions of the brain 
(e.g. the orbitofrontal lobe, for the proximity of the
air surrounding sinuses). This artifact has been
profoundly described, generally within the context of
functional MRI because it is usually acquired with 
\gls*{epi} as \gls*{dmri}.

First approaches to
retrospective correction \cite{jezzard_correction_1995},
relied on an extra \gls*{mri} acquisition (so-called
\emph{field mapping}), that probes the inhomogeneity of the field.
A second theory-based breed of methodologies acquire a 
map of the \gls*{psf} of the \gls*{epi} readouts to correct 
for the artifact \cite{robson_measurement_1997}. 
Next generation of methodologies \cite{cordes_geometric_2000,
chiou_simple_2000}, make use of the acquisition of 
extra \gls*{epi} volumes 
with specific differences that enable correction, for instance
the opposed direction of the phase encoding gradients 
Finally, the last family of methodologies acquire an 
extra T2-weighted image, and use its anatomical correctness 
as reference to find the deformation map through nonlinear 
registration \cite{kybic_unwarping_2000,studholme_accurate_2000}.
Registration-based methods usually map
the T2 image to the so-called baseline (B0) volume 
of \gls*{dmri}. The choice of T2 is due to the strong
similarity of intensity distribution with the B0 of \gls*{dmri}
datasets. More recent works report extensions or combined
approaches of existing techniques
\cite{andersson_how_2003,zaitsev_point_2004,%
holland_efficient_2010,andersson_comprehensive_2012}.

Even though aforementioned techniques for distortion correction
have been studied \cite{zeng_image_2002,wu_comparison_2008},
the lack of a gold-standard limits the possible benchmarking 
strategies. Recently, \cite{irfanoglu_effects_2012} raised the question
of distortion-derived impacts in tractography.
In this work, we propose an evaluation framework
using a digital phantom designed for connectivity assessment.
This framework enabled us to compare several correction 
techniques and characterize their geometrical accuracy 
and the dropout compensation. Finally, we report their 
impact on subsequent tractography and resulting connectivity
matrices.

### 2013-09-28 Publications/2013-ISBI14/03-methods.tex
First review after Manu's comments

\subsection{Digital Phantom}
We modified the digital phantom provided with the
\emph{\gls*{hardi} reconstruction challenge} held
in ISBI 2013, San Francisco, US. This model is
built upon high-resolution \glspl*{tpm}, from which
\gls*{dmri} signal can be simulated with different
\glspl*{snr}. The new version of the phantom includes
a \gls*{csf} distribution, which was not simulated
in the version available for the contest. For this paper,
with these new \glspl*{tpm} we produced \gls*{dti} at
typical low-resolution, and corresponding \gls*{t1} and 
\gls*{t2} images at high-resolution. All the
modalities where numerically simulated, using previously
reported \gls*{mri} typical parameters.

\begin{table}[h]
\caption{Simulated \gls*{mri} modalities}
\label{table:phantom}
\begin{center}
\begin{tabular}{c|c}
\hline
Modality & Description \\
\hline
\gls*{dti} & 30 directions, \\
\hline
\gls*{t1} & \\
\hline
\gls*{t2} & \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Fieldmap-based correction methodologies
\cite{jezzard_correction_1995} use a map
of the field in the scanner. More precisely, the
phase difference between two subsequent samplings
of the fieldmap. With that information, it is possible
to compute the theoretical displacements that each
voxel is to suffer, the so-called \gls*{vsm}. The
most prominent feature of this \gls*{vsm} is that all
the shifts have the same orientation (parallel to the
phase-encoding direction of the \gls*{epi}) and their
magnitude and direction depend on the \gls*{epi} 
gradient increments (or \emph{blips}), and the actual
phase difference at the voxel.

In order to create a realistic distortion, we started
by generating a synthetic phase-difference map 
consistent with the phantom.  We defined two regions 
of smooth phase inhomogeneity and computed the 
corresponding \gls*{vsm} using the tools distributed 
with the \texttt{FSL} package \cite{jenkinson_fsl_2012}.

Then, it is straightforward to use the \gls*{vsm} to 
produce the deformation of the original phantom data,
using the \texttt{fugue} tool included in
\cite{jenkinson_fsl_2012}. With the same procedure,
just setting the opposite encoding direction, we obtained
the corresponding reversed encoding ``acquisition'' of
the same phantom, in order to use it in evaluating the
second family of correction methods.

In this simple manner, we generated a full gold-standard
protocol, containing realistic \gls*{t1} and \gls*{t2}
at high resolution, \gls*{dti} data acquired in two
different phase encoding directions, and the ground-truth
\gls*{dti} data that is not available for real datasets.

>>>>

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the High angular resolution diffusion imaging 
reconstruction Challenge held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as a \gls*{dmri}
(1.0mm isotropic, $b$=1200, 32 evenly-distributed 
directions, 1 B0 image) image.
The phantom includes \gls*{wm} fiber bundles, 
\gls*{csf} and \gls*{gm}. Diffusion is modeled by a 
restricted and a hindered compartment, similar to
\cite{assaf_composite_2005}.

@@

Fieldmap-based correction methodologies
\cite{jezzard_correction_1995} use a map
of the field in the scanner. More precisely, the
phase difference between two subsequent samplings
of the fieldmap. With that information, it is possible
to compute the theoretical displacement that each
voxel is to suffer, the so-called \gls*{vsm}. The
most prominent feature of this \gls*{vsm} is that all
the shifts have the same orientation (parallel to the
phase-encoding direction of the \gls*{epi}) and their
magnitude and direction depend on the \gls*{epi} 
gradient increments (or \emph{blips}), and the actual
phase difference at the voxel.

In order to create a realistic distortion, we
generated a synthetic phase-difference map 
consistent with the phantom.  We defined two regions 
of smooth de-phasing (up to $0.05\pi$ rad. excursions) 
and computed the corresponding \gls*{vsm}, using the 
tools distributed with the \texttt{FSL} package 
\cite{jenkinson_fsl_2012}  and
standard parameters ($\Delta$TE=2.46ms for the
fieldmap and \emph{effective dwell time} of 
0.77ms for the \gls*{epi}).

Once a realistic \gls*{vsm} was available, we generated
the corresponding distorted \glspl*{dwi}, in two opposed
phase-gradient encoding directions. The second simulated
``acquisition'' of the same phantom was necessary 
for evaluating the second family of correction methods.

In this simple manner, we generated a full gold-standard
protocol, containing realistic T1 and T2
at high resolution, \glspl*{dwi} acquired in two
different phase encoding directions, and a ground-truth
\gls*{dwi} data, which is not available for real datasets.
